Approximation as Differentiation: Notes

### 2025-05-26 main/category-of-families.tex
Minor tweaks.

\begin{definition}[Finite dimensional vector space over $\RR$]
Up to isomorphism, an \emph{$n$-dimensional vector space over $\RR$} is the set $\RR^n$ equipped with
component-wise addition of vectors $+: \RR^n \times \RR^n \to \RR^n$ and scalar multiplication $\mult: \RR
\times \RR^n \to \RR^n$ (overloading $+$ and $\mult$), where $(\RR^n,+)$ is an abelian group (with identity
$0$ and additive inverse $-v$ again defined component-wise) and the vector operations are compatible with the
field operations in that the following equations hold:

@@

\begin{definition}[Category $\FinVect$]
Define $\FinVect$ to be the category which has as objects $\RR^n$ all finite-dimensional vector spaces, and as
morphisms $f: \RR^m \to \RR^n$ all functions $f$ satisfying $f(u + v) = f(u) + f(v)$ and $f(a \mult v) = a
\mult f(v)$.
\end{definition}

@@

\vspace{1mm}
\noindent Composition is with bilinear. The direct sum $V \biprod W \iso \RR^{m + n}$ given by the Cartesian
product is a biproduct $(V \biprod W, \biinj_{V}, \biinj_{W}, \pi_1, \pi_2)$ with $\biinj_{V} =
\prodM{\id_{V}}{0_{V,W}}$ and $\biinj_{W} = \prodM{0_{V,W}}{\id_{W}}$.

@@

Derivatives respect composition~\cite{spivak65}. Suppose $f: \RR^m \to \RR^n$ and $g: \RR^n \to \RR^k$. For
any $x \in \RR^m$ we have:

@@

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the constant family of tangent spaces at $\RR^n$ (also written $\RR^n$,
for convenience), and each map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as the pair $(f_1, f_2)$.
Then the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp
\partial f)$ corresponds exactly to the forward chain rule, since $(\reindex{\partial g}{f} \comp \partial
f)_x = {\partial g}_{f(x)} \comp {\partial f}_x$, with the reindexing along $f$ selecting the appropriate
pushforward by recomputing $f(x)$.

>>>>

\begin{definition}[Finite dimensional vector space over $\RR$]
Up to isomorphism, an \emph{$n$-dimensional vector space over $\RR$} is the set $\RR^n$ equipped with
component-wise addition of vectors $+: \RR^n \times \RR^n \to \RR^n$ and scalar multiplication $\mult: \RR
\times \RR^n \to \RR^n$ (overloading $+$ and $\mult$), where $(\RR^n,+)$ is an abelian group (with identity
$0$ and additive inverse $-v$ again defined component-wise), and where the vector operations are compatible
with the field operations in that the following equations hold:

@@

\begin{definition}[Category $\FinVect$]
Define $\FinVect$ to be the category which has as objects $\RR^n$ all finite-dimensional vector spaces, and as
morphisms $f: \RR^m \to \RR^n$ all functions $f$ satisfying $f(u + v) = f(u) + f(v)$ and $f(a \mult v) = a
\mult f(v)$, with $f$ preserving the abelian group structure automatically.
\end{definition}

@@

\vspace{1mm}
\noindent Composition is with bilinear. The direct sum $V \biprod W \iso \RR^{m + n}$ given by the Cartesian
product is a biproduct $(V \biprod W, \biinj_{V}, \biinj_{W}, \pi_1, \pi_2)$ with $\biinj_{V} =
\prodM{\id_{V}}{0_{V,W}}$ and $\biinj_{W} = \prodM{0_{V,W}}{\id_{W}}$. \todo{Significance of biproducts}

@@

Derivatives respect composition~\cite{spivak65}. Suppose $f: \RR^m \to \RR^n$ and $g: \RR^n \to \RR^k$. Then
for any $x \in \RR^m$ we have:

@@

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the constant family of tangent spaces at $\RR^n$ (also written $\RR^n$,
for convenience), and each map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as the pair $(f_1, f_2)$.
Then the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp
\partial f)$ again corresponds exactly to the forward chain rule, since $(\reindex{\partial g}{f} \comp
\partial f)_x = {\partial g}_{f(x)} \comp {\partial f}_x$, with the reindexing along $f$ recomputing $f(x)$ in
order to select the appropriate pushforward.
